RUPEE: A fast and accurate purely geometric protein structure search

### 2019-01-28 docs/rupee_2.tex
standard
revisions

RUPEE has two modes of operation, fast and top-aligned. 
Fast mode is significantly faster than all other protein structure searches discussed below but at the expensive of accuracy.
Despite this, we will show that the accuracy of RUPEE in fast mode is not far below that of the best available structure searches. 
On the other hand, the accuracy and response times of RUPEE in top-aligned mode are comparable to currently available protein structure searches that are commonly considered fast.

RUPEE stands out as not just another protein structure search, of which there are many. 
RUPEE is the first, to our knowledge, purely geometric protein structure search to achieve results as good as the best available protein structure searches. 
Additionally, it can be argued that the kinds of matches that RUPEE does return have more added value than the current state of the art in that with equal scores it is able to return results not biased toward a structure classification hierarchy such as SCOPe or sequence clusters such as the PDB-90. 
In this regard, RUPEE makes a fundamental contribution to protein structure research that lends itself to being leveraged in existing systems. 
It also provided a path for further research activity in the direction of big data representations of protein structures.

@@

\section*{Related work}

@@

@@

\subsection*{Regions of Torsion Angles}

@@

\subsection*{Linear Encoding of Protein Structure}

@@

\subsection*{Response Times}

Response times to a large degree are a measure of the amount of resources available to an application. 
Response times for RUPEE were gathered from RUPEE running on a fairly old laptop with a 2nd generation Intel\textregistered{} Core\texttrademark{} i7-2720QM Quad-Core CPU and 8 GB of memory. 
With more resources the response times of RUPEE can be further improved since pairwise alignments can be run in parallel. 
For mTM, SSM, and CATHEDRAL, we gathered response times by automating their respective web sites using the Selenium WebDriver API.

@@

The trend of increasing response times for RUPEE top-aligned is a direct result of the pairwise structure comparisons that are performed on the top 2000 results provided by RUPEE fast mode.

@@

One possible drawback of RUPEE is that in both fast and top-aligned modes, it only returns the top 400 results. 
Again, with more resources, this number can be increased, but there still remains a need for some kind of cut-off. 
Nonetheless, in most search tools, having to looking past the first few hundred results usually indicates an ineffective search strategy.
To this end, RUPEE provides filters that can be used for traversing structure space more efficiently. 
For SCOPe, CATH, and ECOD you can instruct the search to only return domains that differ from the query structure at a chosen hierarchy level classification. 
Additionally, for CATH you can filter results based on hierarchy level representatives. 
Since RUPEE does not rely on sequence clusters, these kinds of filters are easy to implement, do not reduce the number of returned results, and allow for the discovery of unexpected structural similarities across classification hierarchies.

@@

As can be observed from considering \figurename~\ref{fig:combined_scoring_ce} and \figurename~\ref{fig:combined_precision} together, there are good domain matches aligned with the SCOPe hierarchy that mTM is able to find that RUPEE does not. 
Conversely, there are good domain matches not aligned with the SCOPe hierarchy that RUPEE finds and mTM does not. 
A comprehensive listing of these difference may find interesting similarities not previously known.

@@

With the growth rate of solved structures deposited in the PDB, the need for a fast and scalable structure search is growing. 
Using run position encoded shingles of residue descriptors combined with min-hashing and LSH, we have shown that RUPEE fast is able to provide good results in seconds running on an Quad-Core laptop.
Currently, RUPEE fast is the fastest available protein structure search providing the demonstrated level of accuracy. 
For RUPEE top-aligned, we have shown that a purely geometric big data approach to protein structure search is able to produce results equal to or better than the current state of the art protein structure searches that variously depend on clustered sequences or protein structure classification hierarchies. 
Moreover, we have shown evidence that suggests the results from RUPEE top-aligned provide more added value by discovering high scoring protein structure matches not necessarily aligned with a particular protein structure classification hierarchy or hiding behind cluster representatives. 
The ability for RUPEE to quickly examine all structures among hundreds of thousands sets it apart as a tool that can be used for discovering previously undetected structural similarities.

\section*{Supporting information}

>>>>

RUPEE has two modes of operation, fast and top-aligned. 
Fast mode is significantly faster than all other protein structure searches discussed below but at the expense of accuracy.
Despite this, we will show that the accuracy of RUPEE in fast mode is not far below that of the best available structure searches. 
On the other hand, the accuracy and response times of RUPEE in top-aligned mode are comparable to currently available protein structure searches that are commonly considered fast.

RUPEE stands out as not just another protein structure search, of which there are many. 
RUPEE is the first, to our knowledge, purely geometric protein structure search to achieve results as good as the best available protein structure searches. 
Being purely geometric, the kinds of matches that RUPEE returns are not biased toward a structure classification hierarchy such as SCOPe or sequence clusters such as the PDB-90.
Moreover, without a dependence on pre-calculated results, it is easy to apply a number of filters to the protein structures RUPEE searches in order to better focus on structures of interest.
As one example, domains not having the same fold classification as the query structure can be searched exclusively to investigate unexpected structural relationships across folds. 
In this regard, RUPEE makes a fundamental contribution to protein structure research that also provides a path for further research in the direction of big data representations of protein structures.

@@

\section*{Related Work}

@@

Of the pairwise alignment tools, CE, FATCAT, TM-align and DALI, we have found TM-align is the fastest while also providing high-quality pairwise structure alignments. 
It is for this reason that we use TM-align for RUPEE in top-aligned mode.

@@

\subsection*{Regions of torsion angles}

@@

\subsection*{Linear encoding of protein structure}

@@

\subsection*{Response times}

Response times to a large degree are a measure of the amount of resources available to an application. 
Response times for RUPEE were gathered from the RUPEE web site running on a single Amazon Web Services (AWS) c5.2xlarge elastic compute (EC2) unit. 
With more resources, RUPEE response times can be further improved since pairwise alignments can be run in parallel. 
For mTM, SSM, and CATHEDRAL, we gathered response times by automating their respective web sites using the Selenium WebDriver API.

@@

The trend of increasing response times for RUPEE top-aligned is a direct result of the pairwise structure comparisons that are performed on the top 8000 results provided by RUPEE fast mode.

@@

One possible drawback of RUPEE is that in both fast and top-aligned modes, it only returns the top 400 results. 
Again, with more resources, this number can be increased, but there still remains a need for some kind of cut-off. 
Nonetheless, in most search tools, having to looking past the first few hundred results usually indicates an ineffective search strategy.
To this end, RUPEE provides filters that can be used for traversing structure space more efficiently. 
For SCOPe, CATH, and ECOD you can instruct the search to only return domains that differ from the query structure at a chosen hierarchy level classification. 
Additionally, for CATH you can filter results based on hierarchy level representatives. 
Since RUPEE does not rely on sequence clusters or pre-calculated results, these kinds of filters are easy to implement, do not reduce the number of returned results, and allow for the discovery of unexpected structural similarities across classification hierarchies.

@@

@@

With the growth rate of solved structures deposited in the PDB, the need for a fast and scalable structure search is growing. 
Using run position encoded shingles of residue descriptors combined with min-hashing and LSH, we have shown that RUPEE fast is able to provide good results in seconds running on a single AWS elastic compute unit.
Currently, RUPEE fast is the fastest available protein structure search providing the demonstrated level of accuracy. 
For RUPEE top-aligned, we have shown that a purely geometric big data approach to protein structure search is able to produce results equal to or better than the current state of the art protein structure searches that variously depend on clustered sequences or pre-calculated results. 
The ability for RUPEE to quickly examine all structures among hundreds of thousands sets it apart as a tool that can be used for discovering previously undetected structural similarities.

\section*{Supporting Information}
